---
title: "Practical Machine Learning Course Final Project"
author: "J Gold"
date: "Tuesday, April 05, 2016"
output: html_document
---

I tried several machine learning methods using the caret package before choosing my
final model in order to predict the classe of the 20 test cases.  Prior to choosing
the models, I ran a correlation between variables in the training dataset (see code
below).  After observing that several variables had correlations of > 80%, I decided
to consider principal components analysis (pca) in model building.  I included
random forest and boosting in the list of potential modelsbecause these have been
found to be top performers in prediction contests.  I also considered linear 
discriminant analysis (lda) because a linear model would provide easier interpre
tability.  Hence, the models I chose to evaluate included: random forest, random
forest with pca, lda, and boosting.  In all models, 10-fold cross-validation was
implemented through the trainControl parameter in caret’s train function.  I chose
random forest for my final model because the accuracy was greatest with that method
(see table 1).

Table 1. Model type with accuracy and OOB error rates.


Model type                  | Accuracy | OOB error rate
----------------------------|----------|---------------
Random forest, mtry=3	    | 0.996	   |  0.33%
Random forest, mtry=4	    | 0.996	   |  0.29%
Random forest, mtry=5	    | 0.996	   |  0.30%
Random forest, mtry=8	    | 0.997	   |  0.24%
Random forest w. pca	    | 0.983	   |  1.77%
Linear discriminant analysis| 0.550	   |  -
Stochastic gradient boosting| 0.963	   |  -


In order to tune the random forest model, I used different values of mtry (number of
predictors available for splitting per node).  Note that it was not possible for me
to let the train function choose mtry as it crashed my computer, although this may
have been more optimal.  I chose the mtry value with the lowest out-of-bag (OOB)
error rate, and the greatest accuracy.  I tried mtry values between 3-8.  I stopped
at mtry = 8 because I judged that the accuracy for that model was sufficiently high.
Hence, my final model was random forest with mtry = 8. This had an accuracy of 
0.997, and with an OOB error rate of 0.24%.  Note that the model metrics are shown 
below for the final model, while only the code is shown for the other models that I
tried.  My final model yielded 100% prediction accuracy when running on the test 
set.


```{r setup, include=FALSE}
# enable the cache
knitr::opts_chunk$set(cache=TRUE)
```

```{r}
## remove any global environment variables to save space
rm(list=ls())
## increase memory limit
memory.limit(90000)
## load required libraries
library(caret)
library(randomForest)
library(doParallel)
library(foreach)
library(doSNOW)
library(gbm)
library(survival)
library(HiDimDA)
library(plyr)

## set a seed so results can be duplicated
set.seed(3465)

## read in training and test datasets
datTrain = read.csv("pml-training.csv")
datTest = read.csv("pml-testing.csv")

## eliminate variables not to be used in the model.  Keep the following predictor
# variables: for belt, arm, forearm, dumbbell: pitch, yaw, roll,total acceleration, 
# gyros_xx_x,_y, _z, accel_xx_x, _y, _z, and magnet_xx_x, _y, & _z.  Also, keep
# user_name as one would expect correlation of observations within a particular
## participant.  Keep classe as the outcome.

initDataset <- function(dat) {
    
    ## eliminate variables not to be used in the model
    namedTrain <- names(dat)
    
    gkurtosis <- grep("kurtosis",namedTrain,value=TRUE)
    gskewness <- grep("skewness",namedTrain,value=TRUE)
    gmax <- grep("max_",namedTrain,value=TRUE)
    gmin <- grep("min_",namedTrain,value=TRUE)
    gamplitude <- grep("amplitude_",namedTrain,value=TRUE)
    gstddev <- grep("stddev_",namedTrain,value=TRUE)
    gavg <- grep("avg_",namedTrain,value=TRUE)
    gvar <- grep("var_",namedTrain,value=TRUE)
    gtimestamp <- grep("timestamp",namedTrain,value=TRUE)
    gwindow <- grep("window",namedTrain,value=TRUE)
    
    gexclude <- c(gkurtosis,gskewness,gmax,gmin,gamplitude,gvar,gavg,gstddev,
                  gtimestamp,gwindow,"user_name", "X")
    myvars <- names(dat) %in% c(gexclude)
    ssdatTrain <- dat[!myvars]
    return(ssdatTrain)
}

## only include observations where new_window columns say "no" indicating no summary 
## statistics in training dataset.
tempdatTrain <- subset(datTrain,new_window == "no")
## initialize training dataset
tempTrain <- initDataset(tempdatTrain)
## initialize testing dataset and remove "problem_id" column since we don't
## want "problem_id" to be a predictor variable.
tempTest <-initDataset(datTest)
tempTest <- tempTest[,-54]

## remove variables no longer needed.
rm(tempdatTrain)
rm(datTest)
rm(datTrain)
```
```{r, eval = FALSE}
## Just to get an idea of the dataset, determine correlations among 
# variables.  
M <- abs(cor(tempTrain[,-53]))
diag(M) <- 0
# all variables corr .80 or >
which(M > 0.8,arr.ind=T)   
   
# Since so many variables are correlated, consider factor analysis for dimension 
# reduction (see below).
``` 
```{r}
# set method to cross-validation, with 10 folds
train_control <- trainControl(method="cv",number=10)

# run models over multiple cores to save time
registerDoSNOW(makeCluster(detectCores()/2))
```
```{r, eval = FALSE}
## because letting random forest algorithm without specifying mtry takes way too
## much time, I have opted to set mtry to specific values.

## try random forest w. mtry = 4
newGrid = expand.grid(mtry = 4)
modFitrf4 <- train((as.factor(classe))~., data=tempTrain, method = "rf", 
                  trControl=train_control, tuneGrid = newGrid)
#predict on test set
prf4 <- predict(modFitrf4,tempTest)

## try random forest w. mtry = 3
newGrid = expand.grid(mtry = 3)
modFitrf3 <- train((as.factor(classe))~., data=tempTrain, method = "rf", 
                  trControl=train_control, tuneGrid = newGrid)
#predict on test set
prf3 <- predict(modFitrf3,tempTest)

## try random forest w. mtry = 5
newGrid = expand.grid(mtry = 5)
modFitrf5 <- train((as.factor(classe))~., data=tempTrain, method = "rf", 
                   trControl=train_control, tuneGrid = newGrid)
#predict on test set
prf5 <- predict(modFitrf5,tempTest)
```
```{r}
## This is the final model that I chose.
## try random forest w. mtry = 8
newGrid = expand.grid(mtry = 8)
modFitrf8 <- train((as.factor(classe))~., data=tempTrain, method = "rf", 
                   trControl=train_control, tuneGrid = newGrid)
#predict on test set
prf8 <- predict(modFitrf8,tempTest)
# print prediction and model metrics.
prf8
modFitrf8
modFitrf8$finalModel
```
```{r, eval = FALSE}
## try random forest w. pca, mtry unspecified. Performing a pca in the caret package 
## preProc also scales and centers the variables.  
modFitrfpca <- train((as.factor(classe))~., data=tempTrain, method = "rf", 
                      trControl=train_control, preProcess = "pca")

prfpca <- predict(modFitrfpca, tempTest)
# print prediction and model metrics.
prfpca
modFitrfpca
modFitrfpca$finalModel

## try factor-based linear discriminant analysis; q is # factors
newGrid = expand.grid(q = 5)
modFitlda <- train((as.factor(classe))~., data=tempTrain, method = "RFlda", 
                      trControl=train_control, preProc = c("center","scale"),
                      tuneGrid = newGrid)

plda <- predict(modFitrflda,tempTest)
# print prediction and model metrics.
plda
modFitlda
modFitlda$finalModel

## try boosting
modFitb <- train((as.factor(classe))~., data=tempTrain, method = "gbm", 
                 trControl=train_control, verbose = FALSE)

prb <- predict(modFitb,tempTest)
prb
modFitb
modFitb$finalModel
```


